Type theory in type theory with single substitutions

### 2024-09-14 types_2024_postproc/p.tex
sutff

@@

In Section \ref{sec:ssc}, we introduce the single substitution
calculus only relying on a working knowledge of an implementation of
type theory such as Agda, no prior experience with the metatheory of
type theory is required. We first show how to get rid of explicit
substitutions from the syntax using the technique of
$\alpha$-normalisation (Section \ref{sec:alpha}), and we use this to
prove several admissible equations. Then we show how to derive the
rules of a CwF from our syntax (Section \ref{sec:cwf}). Finally, we
show that our syntax is isomorphic to the CwF-style syntax (Section
\ref{sec:iso}). We conclude in Section \ref{sec:conclusion}.

@@

In the 1990s single substitution calculi for type theory were popular,
but usually in the extrinsic setting, see e.g.\ the thesis of
Altenkirch \cite{alti:phd93}. B-systems were introduced by Voevodsky
\cite{voevodsky2014bsystems} as an algebraic way of describing type
theories close to the notations using typing judgements. Using our
notations B-systems can be described as follows. B-frame.

There are several ways to define type theory as a GAT. For example,
Thomas Ehrhard's calculus \cite{ehrhard,coquandEhrhard}, Dybjer's
categories with families (CwFs
\cite{DBLP:conf/types/Dybjer95,Castellan2021}), contextual categories
or C-systems
\cite{DBLP:journals/apal/Cartmell86,DBLP:journals/lmcs/AhrensLV18},
natural models \cite{DBLP:journals/mscs/Awodey18}, B-systems
\cite{AHRENS_EMMENEGGER_NORTH_RIJKE_2023}.  With the exception of
B-systems, the listed notions all define a parallel substitution
calculus, where contexts and substitutions form a category.

Ehrhard

@@

contextual sCwF and contextual SSC with fct types

Formalisations of type theory use low-level

Algebraic formalisations of type theory: Brunerie

Initiality conjecture.

\cite{initiality-agda} \cite{DBLP:conf/popl/AltenkirchK16},
normalisation, shallow embedding canonicity, gluing. Low level
formalisations: Abel, Loic, metacoq. They can be seen as unfolding the
more abstract thing in a (weak) setoid model.

Other SSC: 90s French, Thorsten thesis, B-systems, Luksa paper.

@@

In a type-theoretic metatheory,
the initial model of a GAT is a quotient inductive-inductive type
(QIIT \cite{DBLP:journals/pacmpl/KaposiKA19}). QIITs are a general
family of inductive types where multiple types are defined mutually,
later types can be indexed over previous ones, and equality
constructors are also allowed, generalising quotients. The elimination
principle of the QIIT corresponds to a structural induction principle
and is equivalent to the statement that the syntax is the initial
model \cite{DBLP:journals/pacmpl/KaposiKA19}, that is, there is a
unique morphism from the syntax to any other model.

@@

@@

>>>>

\usepackage{tikz}
\usetikzlibrary{arrows}

@@

TODO. In Section \ref{sec:ssc}, we introduce the single substitution
calculus only relying on a working knowledge of an implementation of
type theory such as Agda, no prior experience with the metatheory of
type theory is required. We first show how to get rid of explicit
substitutions from the syntax using the technique of
$\alpha$-normalisation (Section \ref{sec:alpha}), and we use this to
prove several admissible equations. Then we show how to derive the
rules of a CwF from our syntax (Section \ref{sec:cwf}). Finally, we
show that our syntax is isomorphic to the CwF-style syntax (Section
\ref{sec:iso}). We conclude in Section \ref{sec:conclusion}.

@@

In the 1990s single substitution calculi for type theory were popular,
but usually in the extrinsic setting, see e.g.\ the thesis of
Altenkirch \cite{alti:phd93}. B-systems were introduced by Voevodsky
\cite{voevodsky2014bsystems} as an algebraic way of describing type
theories close to the notations using typing judgements. B-systems are
a generalised algebraic (intrinsic) presentation of type theory using
single substitutions. xxx

\begin{tikzpicture}
  \node (b0) at (0,0) {$\top\cong B_0$};
  \node (b1) at (2,0) {$B_1$};
  \node (b2) at (4,0) {$B_2$};
  \node (b3) at (6,0) {$\dots$};
  \node (c1) at (3,1) {$\tilde{B}_1$};
  \node (c2) at (5,1) {$\tilde{B}_1$};
  \draw[->,font=\scriptsize] (b1) edge node[below] {$\mathsf{ft}_0$} (b0);
  \draw[->,font=\scriptsize] (b2) edge node[below] {$\mathsf{ft}_1$} (b1);
  \draw[->,font=\scriptsize] (b3) edge node[below] {$\mathsf{ft}_2$} (b2);
  \draw[->,font=\scriptsize] (c1) edge node[above] {$\partial_1$} (b1);
  \draw[->,font=\scriptsize] (c2) edge node[above] {$\partial_2$} (b2);
\end{tikzpicture}

Using our
notations B-systems can be described as follows. B-frame. TODO

@@

The first generalised algebraic presentation of type theory was Thomas
Ehrhard's calculus \cite{ehrhard,coquandEhrhard} which featured a
parallel substitution calculus with $\p$, $\blank^+$ and
$\langle\blank\rangle$ operations, just like in our calculus. Our
single substitution calculus is essentially Ehrhard's calculus with
the categorical composition and identity operations
removed. Categories with families (CwFs
\cite{DBLP:conf/types/Dybjer95,Castellan2021}) feature $\p$, $\q$,
$\blank,\blank$ operations which makes it more apparent that
substitutions are simple lists of terms. CwFs are equivalent to
Ehrhard's calculus, and so are contextual categories/C-systems
\cite{DBLP:journals/apal/Cartmell86,DBLP:journals/lmcs/AhrensLV18},
the natural models of Awodey \cite{DBLP:journals/mscs/Awodey18} and
B-systems \cite{AHRENS_EMMENEGGER_NORTH_RIJKE_2023}.

Kaposi and Luksa \cite{} defined a telescopic single substitution
calculus for simple type theory, it can be seen as a simply typed
version of B-systems. They show that the category of contextual models
of their calculus is equivalent to the category of contextual simply
typed CwFs with function space. Their equivalence result holds even
without the presence of function space. In the presence of function
space, the same equivalence holds for the simply typed version of our
substitution calculus, see the formalisation accompanying this paper.

Most formalisations of type theory are extrinsic
\cite{pujet,abel,metacoq}: they define the syntax as abstract syntax
trees, and equip it with typing and conversion relations. A higher
level representation is working in setoid hell
\cite{chapman09eatitself}: terms are intrinsically well-typed, but
conversion is still an explicit relation. For formalising the
GAT-level syntax, one needs a stronger metatheory than ordinary Agda
or Coq: the metatheory has to support quotient inductive-inductive
types (QIITs \cite{DBLP:journals/pacmpl/KaposiKA19}), in other words,
initial models of GATs. Altenkirch and Kaposi
\cite{DBLP:conf/popl/AltenkirchK16} formalised the syntax of type
theory using postulated QIITs in Agda, together with parametricity and
normalisation \cite{lmcs:4005} proofs. Brunerie and de Boer
\cite{initiality-agda} constructed the initial contextual category in
Agda using postulated quotients. QIITs are supported by the cubical
set model \cite{} and the setoid model \cite{kaposi-qiit-setoid}, thus
Cubical Agda \cite{} and setoid/observational type theory
\cite{setoid,loicThesis} support QIITs. For example, in Cubical Agda,
set-truncated and groupoid-truncated syntaxes of type theory have been
formalised \cite{cohtt}. Without QIITs, tricks such as shallow
embedding the syntax can be used to formalise metatheoretic results
about type theory such as gluing
\cite{kaposi_et_al:LIPIcs:2019:10532}, and its special cases
canonicity and parametricity \cite{kaposi-shallow}.

At the level of abstraction of SOGATs, the difference between single
and parallel substitution calculi are not visible. However, the
metatheory of type theory can be still investigated in this abstract
setting, via the methods of Synthethic Tait Computability
\cite{DBLP:phd/us/Sterling22} or internal sconing
\cite{DBLP:conf/fscd/BocquetKS23}.

@@

@@

no need for parallel substitutions, empty substitution, parallel
weakenings, telescopes, or combinations of these.

say that we assume the QIIT version of this.

@@

termification in the shallow paper is shown equationally complete
